---
title: "Statistical Analysis Script"
author: "Hershi Aytoda"
date: "2026-09-23"
output: pdf_document
---


## Statistical Analysis Script


# Project: Time Since Release and Review Sentiment in Selected Marvel Film Reviews

# Purpose:

# This script performs the statistical analyses used to investigate whether sentiment 
# in selected Marvel film reviews is associated with the time between a film's release and the
# date the review was posted.


# Secondary analyses examine whether average review sentiment differs among the selected films.


# Input:

# marvel_movie_reviews_lexicon_sentiment.csv

# 

# Output:

# - Descriptive summaries

# - Correlation test results

# - Linear regression results and diagnostics

# - One-way ANOVA results

# - Tukey HSD post-hoc comparisons

# - Kruskal-Wallis test

# - Statistical diagnostic plots

# 

# Execution:

# This script should be run after the review dataset has been

# collected and cleaned and the polarity scores have been

# calculated.

# 

# Software:

# R / RStudio

# 

# Required packages: dplyr, ggplot2,car


```{r}
# 1. Load Packages and Data

# dplyr is used for data summaries and grouping observations.
library(dplyr)

# ggplot2 is used to create the visualizations used to
# examine relationships and distributions in the data.
library(ggplot2)

# car is available if we want to run Levene's test for the movie groups.
library(car)

# Find the repository root whether this file is run from the root or SCRIPTS/.
project_root <- normalizePath(getwd())

while (!dir.exists(file.path(project_root, "DATA"))) {
  parent_directory <- dirname(project_root)
  if (identical(parent_directory, project_root)) {
    stop("Could not find the project DATA/ directory.")
  }
  project_root <- parent_directory
}

# Read the repository's processed review dataset.
data_path <- file.path(
  project_root,
  "DATA",
  "marvel_movie_reviews_lexicon_sentiment.csv"
)
data <- read.csv(data_path)

# Display the structure of the dataset to verify that the
# variables were imported with the expected data types.
glimpse(data)



```


```{r}
# 2. Check Variables Used in Statistical Analysis


# Summarize the time since release and polarity score variables
# to check their ranges and identify any unusual values.
summary(data$days_since_release)
summary(data$polarity_score)

# Count missing observations in the two primary variables.
# Missing values would prevent those observations from being
# included in analyses involving both variables.
sum(is.na(data$days_since_release))
sum(is.na(data$polarity_score))

# Check the number of reviews available for each movie.
# This is important because the statistical comparisons
# between movies depend on the number of observations in each group.
table(data$movie)
```

```{r}

# 3. Exploratory Analysis: Sentiment and Time Since Release

# Create a scatterplot to visually examine the relationship  between the number 
# of days since release and review polarity.
#
# A LOESS curve is added to assess whether the relationship
# appears approximately linear or whether there may be a nonlinear pattern.

ggplot(data, aes(
  x = days_since_release,
  y = polarity_score
)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Review Sentiment vs. Time Since Film Release",
    x = "Days Since Release",
    y = "Polarity Score"
  ) +
  theme_minimal()
```

```{r}

# 4. Spearman Rank Correlation

# The primary research question asks whether sentiment is associated with the amount 
# of time between a film's release and the date its review was posted.
#
# Spearman's rank correlation is used because it evaluates whether the variables 
# have a monotonic association without requiring the relationship 
# to be normally distributed.

# The test returns:
# - rho: direction and strength of the association
# - p-value: evidence against the null hypothesis of no
#   monotonic association
cor_test <- cor.test(
  data$days_since_release,
  data$polarity_score,
  method = "spearman",
  exact = FALSE
)

cor_test

```




```{r}
# 5. Spearman Rank Correlation for Reviews With at Least 10 Words

# Very short reviews may provide limited information about the reviewer's overall sentiment.
# This analysis therefore restricts the data to reviews containing at least 10 words.

# The purpose is to determine whether the relationship between
# time since release and polarity score changes when only
# reviews with more substantial text are included.

# Create a filtered dataset containing only reviews with at least 10 words.
data_10_words <- data %>%
  filter(total_word_count >= 10)

nrow(data_10_words)

summary(data_10_words$total_word_count)

# Calculate Spearman's rank correlation between days since release and polarity score.

# Spearman's correlation is used because it measures whether
# the variables have a monotonic relationship and does not
# require the relationship to be normally distributed.

cor_test_10_words <- cor.test(
  data_10_words$days_since_release,
  data_10_words$polarity_score,
  method = "spearman",
  exact = FALSE
)

# Display the correlation coefficient, p-value,
# and other test information.
cor_test_10_words


```

```{r}


# 6. Kruskal-Wallis Test Across Timing Windows


# The previous analyses treated days since release as a
# continuous variable. This analysis instead groups reviews
# according to how long after the film's release the review
# was posted.
#
# Four timing windows are used:
#   <= 1 month   = 30 days or less
#   1 mo-1 yr    = 31 through 365 days
#   1-5 yrs      = 366 through 1,825 days
#   5+ yrs       = more than 1,825 days
#
# The Kruskal-Wallis test is used to determine whether the
# distribution of polarity scores differs across these
# independent timing groups.
#
# This is a nonparametric alternative to a one-way ANOVA and
# does not require polarity scores to be normally distributed
# within each timing group.


# Create a categorical variable that assigns each review
# to one of the four timing windows based on the number
# of days since the film was released.

data <- data %>%
  mutate(
    timing_window = case_when(
      days_since_release <= 30 ~ "<=1 month",
      days_since_release <= 365 ~ "1 mo-1 yr",
      days_since_release <= 1825 ~ "1-5 yrs",
      days_since_release > 1825 ~ "5+ yrs"
    )
  )


# Convert timing_window to a factor and specify the order
# of the groups so that they appear chronologically in
# tables and visualizations.

data$timing_window <- factor(
  data$timing_window,
  levels = c(
    "<=1 month",
    "1 mo-1 yr",
    "1-5 yrs",
    "5+ yrs"
  )
)


# Count the number of reviews in each timing window.
# This confirms that the observations were assigned to
# the intended groups.

table(data$timing_window)


# Calculate descriptive statistics for polarity score
# within each timing window.
#
# The mean and median are reported to describe the center
# of the polarity scores in each group.
#
# The standard deviation describes the amount of variation
# in polarity scores within each timing window.

timing_summary <- data %>%
  group_by(timing_window) %>%
  summarise(
    n = n(),
    mean_polarity = mean(polarity_score, na.rm = TRUE),
    median_polarity = median(polarity_score, na.rm = TRUE),
    sd_polarity = sd(polarity_score, na.rm = TRUE)
  )

timing_summary

# Create a boxplot to compare the distribution of polarity
# scores across the four timing windows.
#
# The boxplot displays the median, spread, and potential
# outliers for each timing group.

ggplot(data, aes(
  x = timing_window,
  y = polarity_score
)) +
  geom_boxplot() +
  labs(
    title = "Review Sentiment Across Timing Windows",
    x = "Time Since Film Release",
    y = "Polarity Score"
  ) +
  theme_minimal()


# Perform the Kruskal-Wallis test.
#
# Null hypothesis:
# The distribution of polarity scores is the same across
# all four timing windows.
#
# Alternative hypothesis:
# At least one timing window has a different distribution
# of polarity scores.

kruskal_test <- kruskal.test(
  polarity_score ~ timing_window,
  data = data
)

# Display the Kruskal-Wallis test results.

kruskal_test
```
```{r}
# 7. Post-Hoc Comparisons for Timing Windows

# If the Kruskal-Wallis test is statistically significant,
# pairwise Wilcoxon rank-sum tests are used to determine
# which timing windows differ from one another.
#
# Holm's adjustment is applied to the p-values to account
# for the fact that multiple pairwise comparisons are being made.

pairwise_wilcox <- pairwise.wilcox.test(
  data$polarity_score,
  data$timing_window,
  p.adjust.method = "holm"
)

# Display the pairwise comparison results.
pairwise_wilcox


```



```{r}

# Quick look at the relationship before fitting the regression models.


ggplot(data, aes(x=days_since_release, y=polarity_score)) + geom_point()


```



```{r}

# Linear regression
# This estimates how polarity changes with each additional day since release.


model_linear <- lm(
  polarity_score ~ days_since_release,
  data = data
)

summary(model_linear)


ggplot(data, aes(x = days_since_release,
                 y = polarity_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Linear Regression of Review Sentiment on Time Since Release",
    x = "Days Since Release",
    y = "Polarity Score"
  ) +
  theme_minimal()


```



```{r}

# Check the linear regression assumptions.
# These plots cover:
# 1. Linearity 
# 2. Constant variance of residuals 
# 3. Normality of residuals 
# 4. Potential influential observations

par(mfrow = c(2, 2))
plot(model_linear)

```



```{r}
# Exponential regression
# This checks whether the relationship changes at a different rate over time.

# Before fitting the model, days_since_release is centered by 
# subtracting the minimum observed value.

# The exponential model has the form:
# polarity_score = a * exp(-b * days_centered) + c 

# Here, a is the starting magnitude and b is the rate of change.

data <- data %>%
  mutate(
    days_centered = days_since_release -
      min(days_since_release)
  )

model_exp <- nls(
  polarity_score ~ a * exp(-b * days_centered) + c,
  data = data,
  start = list(
    a = 0.1,
    b = 0.001,
    c = 0
  )
)

summary(model_exp)

```



```{r}

# Plot the data with the fitted exponential curve.

ggplot(data, aes(
  x = days_since_release,
  y = polarity_score
)) +
  geom_point(alpha = 0.4) +
  stat_function(
    fun = function(x) {
      predict(
        model_exp,
        newdata = data.frame(
          days_centered = x - min(data$days_since_release)
        )
      )
    }
  ) +
  labs(
    title = "Exponential Model of Review Sentiment Over Time",
    x = "Days Since Release",
    y = "Polarity Score"
  ) +
  theme_minimal()

```


```{r}
# Quadratic regression
# This checks for a curved relationship instead of a straight line.
# The model includes:

# - days_since_release 
# - days_since_release squared 

# The squared term lets the fitted line bend upward or downward.





model_quad <- lm(
  polarity_score ~ days_since_release +
    I(days_since_release^2),
  data = data
)

summary(model_quad)

```



```{r}
# Plot the data with the fitted quadratic curve and confidence interval.



ggplot(data, aes(
  x = days_since_release,
  y = polarity_score
)) +
  geom_point(alpha = 0.4) +
  geom_smooth(
    method = "lm",
    formula = y ~ x + I(x^2),
    se = TRUE
  ) +
  labs(
    title = "Quadratic Regression of Review Sentiment on Time Since Release",
    x = "Days Since Release",
    y = "Polarity Score"
  ) +
  theme_minimal()



```



```{r}

# Compare the regression models.

# AIC balances model fit with model complexity. Lower AIC values indicate a model with a
# better fit-complexity tradeoff among the models being compared. 
# The models being compared should be interpreted as alternative descriptions of the same response variable.


# RMSE measures the typical size of the prediction errors.

# It is calculated by taking the square root of the average 
# squared difference between the observed and predicted values. 

# Lower RMSE indicates that the fitted values are, on average, closer to the observed polarity scores.


AIC(model_linear, model_quad)

AIC(model_linear, model_exp)

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

rmse(
  data$polarity_score,
  fitted(model_linear)
)

rmse(
  data$polarity_score,
  fitted(model_exp)
)
```



```{r}

# One-way ANOVA: differences among movies
# This checks whether average polarity differs by movie.

# Null hypothesis: All movie population means are equal. 
# Alternative hypothesis: At least one movie has a different population mean polarity score. 

# Polarity score is the response and movie is the group.

# One-way ANOVA
anova_model <- aov(
  polarity_score ~ movie,
  data = data
)

summary(anova_model)

library(dplyr)

movie_means <- data %>%
  group_by(movie) %>%
  summarise(
    mean_polarity = mean(polarity_score, na.rm = TRUE),
    n = n()
  )

movie_means

TukeyHSD(anova_model)

```



```{r}
# 11. CALCULATE MEAN SENTIMENT BY MOVIE

# Calculate the average polarity score for each movie. 
# The number of reviews for each movie is also reported so that 
# the mean sentiment can be interpreted alongside the sample size.


# Is review length associated with sentiment polarity?

cor.test(
  data$total_word_count,
  data$polarity_score,
  method = "spearman",
  exact = FALSE
)

```
